Model Comparisons

### 2015-11-26 models/Model_Comparison.Rmd
not normalizing scales before calling RSA

# RSA functionality
```{r}
# speaker informativity
# ---------------------
speaker.lhd = function(d, alpha, cost = 0) {
  exp(alpha*(log(d) - cost))
}
# speaker likelihood
# ------------------
speaker.prob = function(rating, degree, m, alpha) {
  numerator = speaker.lhd(m[rating, degree], alpha)
  norm = 0
  for (i in 1:ncol(m)) {
    norm = norm + speaker.lhd(m[rating, i], alpha)
  }
  return(numerator / norm)
}
# non-normalized posterior
# -----------------------
nn.post = function(rating, degree, m, alpha, useprior) {
  prior = priors[rating, "prior.p"]
  speaker.prob(rating, degree, m, alpha) * prior
}
# normalized posterior
# --------------------
norm.post = function(rating, degree, m, alpha, useprior) {
  nn = nn.post(rating, degree, m, alpha, useprior)
  norm = 0
  for (i in 1:5) {
    # summation over all scale mates
    norm = norm + nn.post(i, degree, m, alpha, useprior)
  }
  return(nn / norm)
}  
```

# Run model functionality
```{r}
# run.partial()
# ------------
# Run RSA with model1 (entailment) and model2 (entailment + generic)
run.partial = function(d, alpha = 1, useprior = F, usenone = F) {
  mat = d %>%
    select(stars, degree, speaker.p) %>%
    spread(degree, speaker.p) %>%
    mutate(hi = hi / sum(hi), 
           low = low / sum(low)) %>%
    select(hi, low)

@@

# run.full()
# ------------
# Run RSA with model3 (full) with alternatives
run.full = function(d, alpha = 1, useprior = F, usenone = F) {
  mat = d %>%
    select(stars, degree, speaker.p) %>%
    spread(degree, speaker.p) %>%
    mutate(hi1 = hi1 / sum(hi1),
           hi2 = hi2 / sum(hi2),
           low1 = low1 / sum(low1),
           low2 = low2 / sum(low2)) %>%
    select(hi1, hi2, low1, low2)

@@

## Run comparisons
```{r run_comparisons, message=FALSE}
# Save performance output
performance.output = data.frame(model=rep(NA, 6), r2=rep(NA, 6))
alphas = rep(NA, 6)

# Model 1 - Entailment, no alpha tuning
# -------------------------------------
alphas[1] = 1
m1.noFit = data.partial %>%
  do(run.partial(., alpha = alphas[1], useprior = T, usenone = F))
performance.output[1, ] = c("M1_noFit", round(cor(m1.noFit$listener.p, m1.noFit$pred)^2, 5))

# Model 1 - Entailment, alpha tuning
# ----------------------------------
alphas[2] = tune.alpha(data.partial)
m1.fit = data.partial %>%
  do(run.partial(., alpha = alphas[2]))
performance.output[2, ] = c("M1_fit", round(cor(m1.fit$listener.p, m1.fit$pred)^2, 5))

# Model 2 - Entailment, generic "None", no alpha tuning
# -----------------------------------------------------
alphas[3] = 1
m2.noFit = data.partial %>%
  do(run.partial(., alpha = alphas[3], usenone = T))
performance.output[3, ] = c("M2_noFit", round(cor(m2.noFit$listener.p, m2.noFit$pred)^2, 5))

# Model 2 - Entailment, generic "None", alpha tuning
# --------------------------------------------------
alphas[4] = tune.alpha(data.partial, usenone = T)
m2.fit = data.partial %>%
  do(run.partial(., alpha = alphas[4], usenone = T))
performance.output[4, ] = c("M2_fit", round(cor(m2.fit$listener.p, m2.fit$pred)^2, 5))

# Model 3 - Full, no alpha tuning
# -------------------------------
alphas[5] = 1
m3.noFit = data.full %>%
  do(run.full(., alpha = alphas[5]))
m3.noFit.matched = m3.noFit[matched.items, ]
m3.noFit.matched$degree = ifelse(m3.noFit.matched$degree == "hi1", "hi", "low")
m3.noFit.matched = cbind(m3.noFit.matched, data.partial$listener.p)
# Check that everything lines up between partial and full
all(m3.noFit.matched$scale == data.partial$scale & m3.noFit.matched$degree == data.partial$degree)
colnames(m3.noFit.matched)[length(colnames(m3.noFit.matched))] = "e6.listener.p"
performance.output[5, ] = c("M3_noFit", round(cor(m3.noFit.matched$e6.listener.p, m3.noFit.matched$pred)^2, 5))

# Model 3 - Full, alpha tuning
# ----------------------------
alphas[6] = tune.alpha(data.full, type = "full", compare.data=data.partial)
m3.fit = data.full %>%
  do(run.full(., alpha = alphas[6]))
m3.fit.matched = m3.fit[matched.items, ]
m3.fit.matched$degree = ifelse(m3.fit.matched$degree == "hi1", "hi", "low")
m3.fit.matched = cbind(m3.fit.matched, data.partial$listener.p)
# Check that everything lines up between partial and full
all(m3.fit.matched$scale == data.partial$scale & m3.fit.matched$degree == data.partial$degree)
colnames(m3.fit.matched)[length(colnames(m3.fit.matched))] = "e6.listener.p"
performance.output[6, ] = c("M3_fit", round(cor(m3.fit.matched$e6.listener.p, m3.fit.matched$pred)^2, 5))
```

@@

Overall model performance: $r^2$. M1 and M2 use literal listener values from exp8. M2 includes a generic "None" defined in terms of stars (1: 1.0, 2:0.0, 3:0.0, 4:0.0, 5:0.0). M3 uses the full set of literal listener alternatives from exp10, however r^2 values and tuning reflects comparisons to exp6 pragmatic listener judgments (not exp11).
```{r overall_perf}
performance.output$model = factor(performance.output$model, levels=c(
  "M1_noFit", "M1_fit", "M2_noFit", "M2_fit", "M3_noFit", "M3_fit"
))
# Populate extra model info
performance.output$r2 = as.numeric(performance.output$r2)
performance.output$alphas = alphas
performance.output$alts = c(rep("entailment only", 2), rep("entailment + None", 2), rep("full", 2))
grid.table(performance.output)

qplot(data=performance.output, x=model, y=performance.output$r2,
      geom="bar", stat="identity",
      main = "Model performance", ylab="r^2", xlab="model") +
  geom_text(aes(label = round(r2, 3), y = r2 + 0.02), size = 3) +
  scale_y_continuous(limits=c(0, 1))
```

>>>>

# RSA functionality
```{r}
# speaker informativity
# ---------------------
speaker.inf = function(d, alpha, cost = 0) {
  exp(alpha*(log(d) - cost))
}
# speaker likelihood
# ------------------
speaker.lhd = function(rating, degree, m, alpha) {
  numerator = speaker.inf(m[rating, degree], alpha)
  normalize = sum(sapply(m[rating, ], function(i) {speaker.inf(i, alpha)}))
  return(numerator / normalize)
}
# non-normalized posterior
# -----------------------
nn.post = function(rating, degree, m, alpha, useprior) {
  prior = priors[rating, "prior.p"]
  return(speaker.lhd(rating, degree, m, alpha) * prior)
}
# normalized posterior
# --------------------
norm.post = function(rating, degree, m, alpha, useprior) {
  nn = nn.post(rating, degree, m, alpha, useprior)
  normalize = sum(unlist(sapply(seq(1, 5), function(i){nn.post(i, degree, m, alpha, useprior)})))
  return(nn / normalize)
}
```

# Run model functionality
```{r}
# run.partial()
# ------------
# Run RSA with model1 (entailment) and model2 (entailment + generic)
run.partial = function(d, alpha = 1, useprior = F, usenone = F) {
  mat = d %>%
    select(stars, degree, speaker.p) %>%
    spread(degree, speaker.p) %>%
#     mutate(hi = hi / sum(hi), 
#            low = low / sum(low)) %>%
    select(hi, low)

@@

# run.full()
# ------------
# Run RSA with model3 (full) with alternatives
run.full = function(d, alpha = 1, useprior = F, usenone = F) {
  mat = d %>%
    select(stars, degree, speaker.p) %>%
    spread(degree, speaker.p) %>%
#     mutate(hi1 = hi1 / sum(hi1),
#            hi2 = hi2 / sum(hi2),
#            low1 = low1 / sum(low1),
#            low2 = low2 / sum(low2)) %>%
    select(hi1, hi2, low1, low2)

@@

## Run comparisons
```{r run_comparisons, message=FALSE}
# Save performance output
performance.output = data.frame(model=rep(NA, 6), cor=rep(NA, 6))
alphas = rep(NA, 6)

# Model 1 - Entailment, no alpha tuning
# -------------------------------------
alphas[1] = 1
m1.noFit = data.partial %>%
  do(run.partial(., alpha = alphas[1], useprior = T, usenone = F))
performance.output[1, ] = c("M1_noFit", round(cor(m1.noFit$listener.p, m1.noFit$pred), 5))

# Model 1 - Entailment, alpha tuning
# ----------------------------------
alphas[2] = tune.alpha(data.partial)
m1.fit = data.partial %>%
  do(run.partial(., alpha = alphas[2]))
performance.output[2, ] = c("M1_fit", round(cor(m1.fit$listener.p, m1.fit$pred), 5))

# Model 2 - Entailment, generic "None", no alpha tuning
# -----------------------------------------------------
alphas[3] = 1
m2.noFit = data.partial %>%
  do(run.partial(., alpha = alphas[3], usenone = T))
performance.output[3, ] = c("M2_noFit", round(cor(m2.noFit$listener.p, m2.noFit$pred), 5))

# Model 2 - Entailment, generic "None", alpha tuning
# --------------------------------------------------
alphas[4] = tune.alpha(data.partial, usenone = T)
m2.fit = data.partial %>%
  do(run.partial(., alpha = alphas[4], usenone = T))
performance.output[4, ] = c("M2_fit", round(cor(m2.fit$listener.p, m2.fit$pred), 5))

# Model 3 - Full, no alpha tuning
# -------------------------------
alphas[5] = 1
m3.noFit = data.full %>%
  do(run.full(., alpha = alphas[5]))
m3.noFit.matched = m3.noFit[matched.items, ]
m3.noFit.matched$degree = ifelse(m3.noFit.matched$degree == "hi1", "hi", "low")
m3.noFit.matched = cbind(m3.noFit.matched, data.partial$listener.p)
# Check that everything lines up between partial and full
all(m3.noFit.matched$scale == data.partial$scale & m3.noFit.matched$degree == data.partial$degree)
colnames(m3.noFit.matched)[length(colnames(m3.noFit.matched))] = "e6.listener.p"
performance.output[5, ] = c("M3_noFit", round(cor(m3.noFit.matched$e6.listener.p, m3.noFit.matched$pred), 5))

# Model 3 - Full, alpha tuning
# ----------------------------
alphas[6] = tune.alpha(data.full, type = "full", compare.data=data.partial)
m3.fit = data.full %>%
  do(run.full(., alpha = alphas[6]))
m3.fit.matched = m3.fit[matched.items, ]
m3.fit.matched$degree = ifelse(m3.fit.matched$degree == "hi1", "hi", "low")
m3.fit.matched = cbind(m3.fit.matched, data.partial$listener.p)
# Check that everything lines up between partial and full
all(m3.fit.matched$scale == data.partial$scale & m3.fit.matched$degree == data.partial$degree)
colnames(m3.fit.matched)[length(colnames(m3.fit.matched))] = "e6.listener.p"
performance.output[6, ] = c("M3_fit", round(cor(m3.fit.matched$e6.listener.p, m3.fit.matched$pred), 5))
```

@@

Overall model performance: $r^2$. M1 and M2 use literal listener values from exp8. M2 includes a generic "None" defined in terms of stars (1: 1.0, 2:0.0, 3:0.0, 4:0.0, 5:0.0). M3 uses the full set of literal listener alternatives from exp10, however r^2 values and tuning reflects comparisons to exp6 pragmatic listener judgments (not exp11).
```{r overall_perf}
performance.output$model = factor(performance.output$model, levels=c(
  "M1_noFit", "M1_fit", "M2_noFit", "M2_fit", "M3_noFit", "M3_fit"
))
# Populate extra model info
performance.output$cor = as.numeric(performance.output$cor)
performance.output$alphas = alphas
performance.output$alts = c(rep("entailment only", 2), rep("entailment + None", 2), rep("full", 2))
grid.table(performance.output)

qplot(data=performance.output, x=model, y=performance.output$cor,
      geom="bar", stat="identity",
      main = "Model performance", ylab="r^2", xlab="model") +
  geom_text(aes(label = round(cor, 3), y = cor + 0.02), size = 3) +
  scale_y_continuous(limits=c(0, 1))
```
